FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2024-12-19 manuscript.Rmd
update 2024-12-19 11:16:18.873357

---
title             : "FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories"
shorttitle        : "FAIR THEORY"

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,4"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "5"
    role:
      - Conceptualization
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
  - id            : "5"
    institution   : "Other affiliations"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`.

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more findable, accessible, interoperable and reusable [REF] for both humans and computers.
Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprecht2020towards; @barker2022introducing], computational workflows [@wilkinson2024applying] or simulation studies [@amaro2024need].
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital representation of a scientific theory, compliant with the FAIR principles.
By improving the efficiency of scholarly communication, FAIR Theory has the potential to foster and accelerate cumulative knowledge acquisition and ultimately advance social scientific research.

@@

## The Need for FAIR Theory

The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely produce hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate findings that are mutually inconsistent,
as the theory's central claims evade falsification.<!--CJ: Resolved AB's comment? AB: Your original wording is contradicting itself; if findings contradict a theory <=> the theory does not explain the findings; I am not sure what the best wording is; should we differentiate between a core of a theory and auxiliary parts?-->
<!-- CJ: Maybe we should give an example here. Does anyone know one? -->

@@

Given these concerns, it is unfortunate that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of *FAIR Theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because deductive research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of reported deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020], and researchers are simply testing them as part of a cultural ritual [@gigerenzerNullRitualWhat2004].
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Or to adapt an analogy of @van2021theory, just collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings.

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' *empirical content* [REF] as it forces researchers to use precise statements, for example by specifying exact functional forms for relations or forces them to specify processes  that would otherwise get away with their vague formulation in verbal theories.
For example, the phonological loop in Baddeley's verbal description of his working memory model allows at least 144 different implementations, for example, depending on how decay rate, recall success, or rehearsal sequence are precisely implemented [@lewandowsky2010computational].
More precise theories makes them easier to falsify,
which necessitates revising them,
thus advancing our principled understanding of the phenomena they describe.
FAIR Theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
It is therefore important for us to emphasize that FAIR Theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down. 
The guidelines introduced by FAIR Theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
Conversely, a directed acyclic graph (DAG) is a type of formal theory,
but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
then this formal theory is not FAIR.
FAIR Theory is thus consistent with, but does not require, formal theory (also see *Accessibility*).

@@

If we consider the current state of Reusability in psychological theory, there appears to be a norm *against* theory reuse:
*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
This norm impedes scientific progress.
Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data [@degrootMethodologieGrondslagenVan1961].
<!-- Theories are made reusable by thorough documentation, -->
<!-- appropriate licensing,  -->
<!-- and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning). -->
In our workshops on FAIR Theory, we similarly notice reluctance to the notion of reusing and adapting theories,
reflected in questions such as "who owns theory",
and "who determines how a theory may be reused or changed"?
These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.

Licensing theories for reuse provides an unambiguous answer to such questions.
In determining what license is appropriate for theory,
it is important to consider that copyright law limits authors' rights based on the *idea-expression dichotomy* [@bently2010copyright],
which holds that copyright explicitly does not
*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*.
Copyright may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
It thus seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
If copyright limits Reusability and does not cover ideas in their purest form (like formal theories),
then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection.
Furthermore, even if copyright would apply, academic research is covered under "fair use" exemptions,
so copyright would pose few restrictions to Reusability in scholarly communication.
Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR Theory;
it explicitly waives all rights and encourages reuse.
In principle, CC0 does not require attribution.
Nevertheless, is essential that scholars do comprehensively cite theory,
including prior work that new theories are based on,
even in absence of legal obligations to do so,
to meet the definition of Reusability (R1.2, Table \@ref(tab:tabfair) and to comply with other definitions of open scholarship [@aalbersbergMakingScienceTransparent2018].

@@

<!-- ## FAIR Theory and Recognition & Rewards -->

@@

## Using FAIR Theory to Perform Causal Inference

@@

One remaining issue that intersects with FAIR Theory is the measurement and operationalization of psychological constructs.
Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":  <!--AB very important point that I was already missing in the introduction; seems we also need FAIR measurements :) -->
it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
Additionally, the "jingle-jangle" fallacy is prevalent in the social sciences:
the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
FAIR Theory can help address the measurement crisis:
since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of

FAIR Theory incorporates theory into open science workflows,
facilitates scholarly communication about theories,
making it easier to share theories with less opportunity for ambiguity and misunderstanding.
FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
removes barriers for knowledge exchange with the global scholarly community,
opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.

>>>>

---
title             : "FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories"
shorttitle        : "FAIR THEORY"

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,4"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Felix Schönbrodt"
    affiliation: "5"
    role:
      - Conceptualization
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
  - id            : "5"
    institution   : "Other affiliations"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`.

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more findable, accessible, interoperable and reusable [REF] for both humans and computers.
Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprecht2020towards; @barker2022introducing], computational workflows [@wilkinson2024applying] or simulation studies [@amaro2024need].
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital representation of a scientific theory, compliant with the FAIR principles.
By improving the efficiency of scholarly communication, FAIR theory has the potential to foster and accelerate cumulative knowledge acquisition and ultimately advance social scientific research.

@@

## The Need for FAIR theory

The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was to blame for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely produce hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate findings that are mutually inconsistent,
as the theory's central claims evade falsification.<!--CJ: Resolved AB's comment? AB: Your original wording is contradicting itself; if findings contradict a theory <=> the theory does not explain the findings; I am not sure what the best wording is; should we differentiate between a core of a theory and auxiliary parts?-->
<!-- CJ: Maybe we should give an example here. Does anyone know one? -->

@@

Given these concerns, it is unfortunate that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of reported deductive research reveals, however, that the link between theory and hypothesis is often tenuous [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit [@friedTheoriesModelsWhat2020, @norouziCapturingCausalClaims2024].
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020], and researchers are simply testing them as part of a cultural ritual [@gigerenzerNullRitualWhat2004].
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Or to adapt an analogy of @van2021theory, just collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings.

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' *empirical content* [REF] as it forces researchers to use precise statements, for example by specifying exact functional forms for relations or forces them to specify processes  that would otherwise get away with their vague formulation in verbal theories.
For example, the phonological loop in Baddeley's verbal description of his working memory model allows at least 144 different implementations, for example, depending on how decay rate, recall success, or rehearsal sequence are precisely implemented [@lewandowsky2010computational].
More precise theories makes them easier to falsify,
which necessitates revising them,
thus advancing our principled understanding of the phenomena they describe.
FAIR theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
It is therefore important for us to emphasize that FAIR theory imposes no restrictions on researchers regarding the manner in which theories are derived and written down. 
The guidelines introduced by FAIR theory primarily pertain to how theories are documented and shared in digital environments, with the aim of enhancing their reusability and extensibility.
For example, it is possible to represent a collection of verbal propositions (perhaps derived through qualitative research) as a FAIR theory.
Conversely, a directed acyclic graph (DAG) is a type of formal theory,
but if it is embedded within a journal article as a bitmap image without any key words to help search engines index that article as a theory paper,
then this formal theory is not FAIR.
FAIR theory is thus consistent with, but does not require, formal theory (also see *Accessibility*).

@@

If we consider the current state of Reusability in psychological theory, there appears to be a norm *against* theory reuse:
*"[Theories are] like toothbrushes — no self-respecting person wants to use anyone else's"* [@mischelToothbrushProblem2008].
This norm impedes scientific progress.
Cumulative knowledge acquisition requires reusable theories that are continuously updated based on insights from new data [@degrootMethodologieGrondslagenVan1961].
<!-- Theories are made reusable by thorough documentation, -->
<!-- appropriate licensing,  -->
<!-- and detailed provenance (e.g., through version control, cross-referencing, and semantic versioning). -->
In our workshops on FAIR theory, we similarly notice reluctance to the notion of reusing and adapting theories,
reflected in questions such as "who owns theory",
and "who determines how a theory may be reused or changed"?
These questions imply a norm against modifying theory without consent from the author reminiscent of the aforementioned problem of dependency on the author.

Licensing theories for reuse provides an unambiguous answer to such questions.
In determining what license is appropriate for theory,
it is important to consider that copyright law limits authors' rights based on the *idea-expression dichotomy* [@bently2010copyright],
which holds that copyright explicitly does not
*"extend to any idea, procedure, process, system, method of operation, concept, principle, or discovery"*.
Copyright may, however, extend to creative works expressing that idea (e.g., writing, visual illustrations).
It thus seems that vague, ambiguous verbal explanations of theories - in other words, those that fall short of the Accessibility criterion - are more likely to qualify for copyright protection than formal theories.
If copyright limits Reusability and does not cover ideas in their purest form (like formal theories),
then it might be counterproductive and possibly misleading to adopt a license that assumes copyright protection.
Furthermore, even if copyright would apply, academic research is covered under "fair use" exemptions,
so copyright would pose few restrictions to Reusability in scholarly communication.
Given these considerations, the CC0 (no rights reserved) license seems most appropriate for FAIR theory;
it explicitly waives all rights and encourages reuse.
In principle, CC0 does not require attribution.
Nevertheless, is essential that scholars do comprehensively cite theory,
including prior work that new theories are based on,
even in absence of legal obligations to do so,
to meet the definition of Reusability (R1.2, Table \@ref(tab:tabfair) and to comply with other definitions of open scholarship [@aalbersbergMakingScienceTransparent2018].

@@

<!-- ## FAIR theory and Recognition & Rewards -->

@@

## Using FAIR theory to Perform Causal Inference

@@

One remaining issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":  <!--AB very important point that I was already missing in the introduction; seems we also need FAIR measurements :) -->
it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
Additionally, the "jingle-jangle" fallacy is prevalent in the social sciences:
the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
FAIR theory can help address the measurement crisis:
since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of

FAIR theory incorporates theory into open science workflows,
facilitates scholarly communication about theories,
making it easier to share theories with less opportunity for ambiguity and misunderstanding.
FAIR Theories are easier to find, and facilitate sharing, reusing, and updating open theories.
More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
removes barriers for knowledge exchange with the global scholarly community,
opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.
